---
title: "Modeling and prediction for movies"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

```{r load-data}
load("movies.Rdata")
```



* * *

## Part 1: Data

The data set is comprised of 651 randomly sampled movies, described by 32 variables.  The sources for the data set are the Rotten Tomatoes and IMDB APIs; URLs to both websites are provided for each movie in the data set.  According to the directions for the course assignment, the movies were produced and released before 2016.  In fact, the movies all had theatrical releases before 2015, although some had DVD releases in 2015.  The range of theatrical releases covered by the sample ranges from the years 1970 to 2014, a span of 45 years.  However, only one movie in the sample has a theatrical release in 1970 and no movies included were released in 1971.

```{r}
movies %>%
  summarize(min(thtr_rel_year), max(thtr_rel_year))
```

A histogram of movie theatrical releases from the sample shows an increase over time.  One may presume that this sampling result reflects the fact that the number of movies produced and released per year has increased over time so that a simple random sample from this time span would end up providing more movies in the sample from later years in the time span than from earlier years.  Exactly how a random sample (simple random sample? stratified sample?) was constructed was not provided by the data set.  The degree of the jaggedness in the histogram is a little puzzling.

```{r}
ggplot(data = movies, aes(x = thtr_rel_year)) + 
  geom_histogram(binwidth = 1)
```


Since random sampling was used, the results are generalizable.  However, given the lack of movies from earlier eras, one should be cautious in extrapolating results to earlier eras of movie making (such as the 1920s, '30s, or '40s).

The data does not derive from an experiment with a randomized control group; so, causal conclusions are not warranted.

* * *

## Part 2: Research question



* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.

* * *

## Part 4: Modeling

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.

* * *

## Part 5: Prediction

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.

* * *

## Part 6: Conclusion

